Minimax-optimal Laplacian Eigenmaps regression over Sobolev Spaces with Neighborhood Graphs

### 2021-03-23 manuscripts/laplacian_eigenmaps/laplacian_eigenmaps_regression.tex
All sections of Laplacian eigenmaps manuscript passed over.

\documentclass{article}
\usepackage{amsmath}
\usepackage{amsfonts, amsthm, amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\hypersetup{
	colorlinks=true,
	linkcolor=blue,
	citecolor=blue
}
\usepackage[parfill]{parskip}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{enumerate}
\usepackage[shortlabels]{enumitem}
\usepackage{fullpage}
\usepackage{mathtools}
\usepackage{tikz}
\usepackage{bm}
\usepackage{multirow}

@@

Suppose we observe data $X_1,\ldots,X_n$ sampled independently from an unknown distribution $P$. As a replacement for $P$, we form a geometric graph $G$ over the observed data, with weighted edges $W_{ij}$ corresponding to proximity between samples $X_i$ and $X_j$. Geometric graphs encode information about $P$--- for instance, the number of connected components of its support, and the shape of each connected component---in an extremely general manner, and can be leveraged to conduct many different fundamental statistical tasks. These include clustering, manifold learning, semi-supervised learning, classification, and regression---both estimation and goodness-of-fit testing. Though much theoretical work has been done on the consistency of learning methods defined with respect to geometric graphs---and more recently, rates of convergence have been established for some methods---little is known so far about their optimality, even for classic statistical tasks.

In this paper we focus on the theoretical statistical properties of \textcolor{red}{graph-based regression}. We assume that in addition to the design points $X_1,\ldots,X_n$ one observes real-valued responses $Y_1,\ldots,Y_n$, and seeks to learn the unknown regression function $f_0(x) := E[Y|X = x]$. The methods we will study are based on \emph{Laplacian eigenmaps}, first introduced by \cite{belkin03a}, which takes the eigenvectors of the graph Laplacian $L$. The Laplacian $L$ is a difference operator, acting on functions $f: \{X_1,\ldots,X_n\} \to \Reals$ as follows,
\begin{equation}
\label{eqn:graph_laplacian}
Lf(X_i) = \sum_{j = 1}^{n} \bigl(f(X_i) - f(X_j)\bigr)W_{ij}. 
\end{equation}
The Laplacian $L$ can be seen as a discretization of the Laplace-Beltrami operator $\Delta_P$, defined as
\begin{equation}
\label{eqn:fokker_planck_1}
\Delta_Pf= \frac{-1}{p} \mathrm{div}(p^2 \nabla f) = \Delta f - 2\frac{\nabla p^{\top} \nabla f}{p}.
\end{equation}
The eigenvectors of $L$ form an orthonormal basis of $L^2(X)$, and are approximations to the eigenvectors of $\Delta_P$. Their corresponding eigenvalues are estimates of~\textcolor{red}{(...)}, and provide a notion of smoothness for each eigenvector---the smaller the eigenvalue, the smoother the corresponding eigenvector.

Laplacian eigenmaps is thus an example of a \emph{spectral series} estimator. A spectral series estimator is a special case of an orthogonal series estimator, one of the most classical methods of non-parametric regression. Traditionally, orthogonal series estimators are defined with respect to a fixed set of basis functions, orthogonalized in $L^2(Q)$ with respect to some pre-determined reference measure $Q$. Spectral series estimators take a particular orthogonal basis, the eigenfunctions of $\Delta_Q$. In contrast, eigenvectors of the graph Laplacian are data-dependent objects, and adapt to the geometry of the design distribution $P$ in a rich manner.

The theoretical properties of (classical) spectral series methods are various, and are in general extremely well-understood \textcolor{red}{(references)}. In particular, they are minimax rate optimal for nonparametric regression over H\"{o}lder and Sobolev spaces. However, in practice these methods suffer from some serious drawbacks. As a practical matter,  even finding the eigenfunctions of $\Delta_Q$ is in general non-trivial. The statistical optimality of such estimators also hinges on the reference measure $Q$ being equal to $P$, an unrealistic assumption when $P$ is unknown. If $Q \neq P$, the standard recommendation is to instead perform nonparametric least squares---this can be statistically rate-optimal, but is numerically unstable when the dimension $d$ is even moderately large.

\textcolor{red}{(TODO)}: Change the preceding paragraph.

Laplacian eigenmaps avoids these drawbacks \textcolor{red}{(...)} On the other hand its statistical properties are not as well understood \textcolor{red}{(...)}. Thus, there exists a theoretical gap.

@@

\begin{quote}
	\textcolor{red}{(TODO)}
\end{quote}

Broadly speaking, we show that when the regression function $f_0$ is smooth, in the Sobolev sense of having weak derivatives bounded in $\Leb^2$-norm, Laplacian eigenmaps methods are statistically minimax optimal, for both estimation and goodness-of-fit testing. This statement holds for different relations between the dimension $d$ and number of derivatives $s$, depending on the problem (estimation or testing).
\begin{table}
	\begin{center}
		\begin{tabular}{p{.2\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Smoothness order & Flat Euclidean (Model~\ref{def:model_flat_euclidean}) & Manifold (Model~\ref{def:model_manifold}) \\
			\hline
			$s \leq 3$ & $\bm{n^{-2s/(2s + d)}}$ & $\bm{n^{-2s/(2s + m)}}$ \\
			$s > 3$  & $\bm{n^{-2s/(2s + d)}}$ & $n^{-6/(6 + m)}$
		\end{tabular}
	\end{center}
	\caption{Summary of estimation rates over Sobolev balls. Bold font marks minimax optimal rates. In each case, rates hold for all $d \in \mathbb{N}$ (under Model~\ref{def:model_flat_euclidean}), and for all $d, m \in \mathbb{N}$ (under Model~\ref{def:model_manifold}). Although we suppress it for simplicity, in all cases the dependence of the error rate on the radius $M$ of the Sobolev ball is also optimal, as long as $M \gtrsim n^{-1/2}$.}
	\label{tbl:estimation_rates}
\end{table}

\begin{table}
	\begin{center}
		\begin{tabular}{p{.175\textwidth} p{.175\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Smoothness order & Dimension & Flat Euclidean (Model~\ref{def:model_flat_euclidean}) & Manifold (Model~\ref{def:model_manifold}) \\
			\hline
			\multirow{2}{*}{$s = 1$} & $\dim(\mc{X}) < 4$ & $\bm{n^{-4s/(4s + d)}}$ & $\bm{n^{-4s/(4s + m)}}$ \\
			& $\dim(\mc{X}) \geq 4$ & $\bm{n^{-1/2}}$ & $\bm{n^{-1/2}}$ \\
			\hline
			\multirow{3}{*}{$s = 2$ or $3$} & $\dim(\mc{X}) \leq 4$  & $\bm{n^{-4s/(4s + d)}}$ & $\bm{n^{-4s/(4s + m)}}$ \\
			& $4 <\dim(\mc{X}) < 4s$  & $n^{-2s/(2(s - 1) + d)}$ & $n^{-2s/(2(s - 1) + m)}$\\
			& $\dim(\mc{X}) \geq 4s$ & $\bm{n^{-1/2}}$ & $\bm{n^{-1/2}}$ \\
			\hline
			\multirow{3}{*}{$s > 3$} & $\dim(\mc{X}) \leq 4$ & $\bm{n^{-4s/(4s + d)}}$ & $n^{-12/(12 + d)}$ \\
			& $4 < \dim(\mc{X}) < 4s$ & $n^{-2s/(2(s - 1) + d)}$ & $n^{-6/(4 + m)}$ \\
			& $\dim(\mc{X}) \geq 4s$ & $\bm{n^{-1/2}}$ & $\bm{n^{-1/2}}$ \\
		\end{tabular}
	\end{center}
	\caption{Summary of testing rates over Sobolev balls. Bold font marks minimax optimal rates. Rates when $d > 4s$ assume that $f_0 \in L^4(P,M)$. Although we suppress it for simplicity, in all cases the dependence of the error rate on the radius $M$ of the Sobolev ball is also optimal, as long as $M \gtrsim n^{-1/2}$.}
	\label{tbl:testing_rates}
\end{table}

\paragraph{Related Work.} 
\begin{itemize}
	\item \textcolor{red}{(TODO)} Summarize what is known on fixed graphs.
	\item \textcolor{red}{von Luxburg, Belkin, Slepcev} show that for a fixed index $k$, the eigenvector $v_k$ converges to the eigenfunction $\psi_k$, in the sense that \textcolor{red}{(...)}. \textcolor{red}{Shi, Burago, Garcia Trillos, Calder, Cheng} build on this, giving finite sample bounds, rates of convergence, and making statements uniform over $1 \leq k \leq k_{\max}$. This suggests that Laplacian eigenmaps may be good alternatives to classical spectral series methods when the latter cannot be implemented (e.g. because $P$ or $\mc{X}$ are unknown.) 
	\item There has been limited analysis of the estimator $\wh{f}$. \cite{zhou2011} consider a similar estimator, but in the semi-supervised setting where the number of unlabeled points grows to infinity. In this case the estimator reduces to a classical spectral series estimator, and there is no error incurred by approximating $\Delta_P$. \cite{lee2016} consider the \emph{diffusion maps} estimator---which uses the eigenvectors of a different normalization of the Laplacian $L$---in both the supervised and semi-supervised setups, but they derive suboptimal rates for the supervised problem. As far as we know, there has been no analysis of the test statistic $\wh{T}$ in the random design framework which we study.
	\item \textcolor{red}{(TODO)}: Alternatively, \cite{trillos2020} \textcolor{blue}{(Green 2021)} use the graph Laplacian operator to induce a penalty over functions $f: \{\bf X\} \to \Reals$. The \emph{Laplacian smoothing} estimator $\wh{f}_{\mathrm{LS}}$ is obtained by minimizing the sum of this penalty along with a data-fidelity term,
	\begin{equation*}
	\wh{f}_{\mathrm{LS}} := \|Y - f\|_n^2 + \lambda \dotp{L_{n,\varepsilon}f}{f}_n.
	\end{equation*}
	\textcolor{blue}{(Green 2021)} show that the resulting estimator is minimax optimal, but only for $s = 1$ and $d \leq 4$. In contrast, the Laplacian eigenmaps estimator is optimal for all $s$ and $d$. 
\end{itemize}

\textbf{(5) Organization.}

\textcolor{red}{(TODO)}

\textbf{(6) Notation.}

@@

\begin{itemize}
	\item Define $\Leb^2(\mc{X}), \Leb^2(P)$ and $\Leb^2(P_n)$. Define $C^k(\mc{X})$, its norm, and its normed ball.
	\item Define asymptotic notation: $\ll$, $\lesssim$ and $\asymp$.
	\item Define min and max notation: $\vee$ and $\wedge$.
\end{itemize}

@@

In the following, recall that the injectivity radius of a $m$-dimensional Riemannian manifold $\mc{X}$ is the maximum value of $\delta$ such that the exponential map $\exp_x: B_m(0,\delta) \subset T_x(\mc{X}) \to B_{\mc{X}}(x,\delta) \subset \mc{X}$ is a diffeomorphism for all $x \in \mc{X}$.
\begin{definition}[Manifold model]
	\label{def:model_manifold}
	The data $(X_1,Y_1),\ldots,(X_n,Y_n)$ are sampled according to~\eqref{eqn:model}. 
	The support $\mc{X}$ of the design distribution $P$ is a closed, connected, smooth and boundaryless Riemannanian manifold embedded in $\Rd$, of intrinsic dimension $1 \leq m < d$. The injectivity radius of $\mc{X}$ is upper bounded by a constant $i_0$. The design distribution $P$ admits a density $p$ with respect to the volume form $d\mu$ induced by the Riemannian structure, which is bounded away from $0$ and $\infty$,
	\begin{equation*}
	0 < p_{\min} \leq p(x) \leq p_{\max} < \infty, \quad \textrm{for all $x \in \mc{X}$.}
	\end{equation*}
\end{definition}

\subsection{Laplacian eigenmaps}
We now formally define the estimator and test statistic we study. Both are derived from eigenvectors of the neighborhood graph Laplacian operator. For a positive, symmetric kernel $\eta: [0,\infty) \to [0,\infty)$, and a bandwidth parameter $\varepsilon > 0$, the \emph{neighborhood graph Laplacian} operator $L_{n,\varepsilon}$ is
\begin{equation}
\label{eqn:neighborhood_graph_laplacian}
L_{n,\varepsilon}u(x) = \frac{1}{n\varepsilon^{2 + \mathrm{dim}(\mc{X})}} \sum_{i = 1}^{n} \bigl(u(x) - u(X_j)\bigr) \eta\biggl(\frac{\|x - X_j\|}{\varepsilon}\biggr)
\end{equation}
(Here $\mathrm{dim}(\mc{X})$ stands for the dimension of $\mc{X}$. It is equal to $d$ under the assumptions of Model~\ref{def:model_flat_euclidean}, and equal to $m$ under the assumptions of Model~\ref{def:model_manifold}. The pre-factor $(n\varepsilon^{2 + \mathrm{dim}(\mc{X})})^{-1}$ is purely for convenience in taking limits as $n \to \infty, \varepsilon \to 0$).

@@

On a different note, it is worth explicitly pointing out that $\wh{f}$ is defined only in-sample, that is, only at the design points $X_1,\ldots,X_n$. There is no straightforward or agreed upon way to extend $\wh{f}$ out-of-sample, and we address the problem of out-of-sample estimation in some detail later in Section~\ref{sec:out_of_sample}.

@@

\paragraph{Flat Euclidean case.}
Take $\mc{X}$ to be an open, connected set with Lipschitz boundary, as in Model~\ref{def:model_flat_euclidean}. Let $Q$ be a measure supported on a subset of $\mc{X}$, that is absolutely continuous with respect to the full-dimensional Lebesgue measure $\nu$. Recall that for a given multiindex $\alpha \in \mathbb{N}^d$, a function $f$ is \emph{$\alpha$ weakly differentiable} if there exists some $h \in L^1(Q)$ such that
\begin{equation*}
\int_{\mc{X}} h g = (-1)^{|\alpha|} \int_{\mc{X}} f D^{\alpha}g, \quad \textrm{for every $g \in C_0^{\infty}(\mc{X})$.}
\end{equation*}
If such a function $h$ exists, it is the $\alpha$th weak partial derivative of $f$, and denoted by $D^{\alpha}f = h$. If $f$ is $\alpha$ weakly differentiable for all $|\alpha| = j$, then $d^jf := \sum_{|\alpha| = j}D^{\alpha}f$ is the sum of all weak partial derivatives of order $j$.

For an integer $s \geq 1$, the Sobolev class $H^s(Q)$ consists of all those functions $f \in \Leb^2(Q)$ which have weak derivatives of all orders $j = 1,\ldots,s$, all of which also belong to $\Leb^2(Q)$. The $j$th order semi-norm for $f \in H^s(Q)$ is $|f|_{H^j(Q)} := \|d^jf\|_{L^2(Q)}$. The corresponding norm
\begin{equation*}
\|f\|_{H^s(Q)}^2 := \|f\|_{\Leb^2(Q)}^2 + \sum_{j = 1}^{s} |f|_{H^j(Q)}^2,
\end{equation*}
which induces the ball
\begin{equation*}
H^s(Q;M) := \bigl\{f \in H^s(Q): \|f\|_{H^s(Q)} \leq M\bigr\}.
\end{equation*} 
In the special case where $Q$ is the uniform measure over $\mc{X}$, we use the notation $H^s(\mc{X}) := H^s(Q)$. Finally, we note that $H^s(Q)$ is the completion of $C^{(\infty)}(\mc{X})$ with respect to the $\|\cdot\|_{H^s(Q)}$ norm, so that $C^{\infty}(\mc{X})$ is dense in $H^s(\mc{X})$.

\paragraph{Manifold case.}

\textcolor{red}{(TODO)}

\paragraph{Boundary conditions.}
In the flat Euclidean case, when $s > 1$ we will assume that $f_0$ satisfies some boundary conditions. In particular, we will assume that $f_0$ belongs to the set of \emph{zero trace} Sobolev functions $H_0^{s}(P)$, defined as the closure of $C_0^s(\mc{X})$ with respect to $\|\cdot\|_{H^s(P)}$ norm. Of course, under Model~\ref{def:model_manifold} the domain $\mc{X}$ is without boundary, and so boundary conditions are irrelevant.

>>>>

\documentclass{article}
\usepackage{amsmath}
\usepackage{amsfonts, amsthm, amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\hypersetup{
	colorlinks=true,
	linkcolor=blue,
	citecolor=blue
}
\usepackage[parfill]{parskip}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{enumerate}
\usepackage[shortlabels]{enumitem}
\usepackage{fullpage}
\usepackage{mathtools}
\usepackage{tikz}
\usepackage{bm}
\usepackage{multirow}
\usepackage[font={small,it}]{caption}

@@

Suppose we observe data $X_1,\ldots,X_n \in \Reals^d$, sampled independently from an unknown distribution $P$. As a replacement for $P$, we form a geometric graph $G$ over the observed data, with vertices at $X_1,\ldots,X_n$ and weighted edges $W_{ij}$ corresponding to proximity between samples $X_i$ and $X_j$. Geometric graphs encode both local information and global information about $P$ in an extremely general manner. For this reason they have been leveraged to conduct many different fundamental statistical tasks, such clustering, manifold learning, semi-supervised learning, classification, and regression \textcolor{red}{(TODO): references}. Substantial theoretical progress has been made regarding the consistency of \emph{graph-based learning}, that is, learning algorithms defined with respect to geometric graphs \textcolor{red}{(TODO): references}. This work sheds light on why such procedures work, by showing that they converge to interesting \emph{continuum} limits as $n \to \infty$. However, thus far little has been said regarding the optimality of graph-based learning methods, even for classic statistical tasks.

In this paper we focus on the theoretical statistical properties of a particularly popular graph-based learning method for regression. We assume that in addition to the design points $X_1,\ldots,X_n$ one observes real-valued responses $Y_1,\ldots,Y_n$, and seeks to learn the unknown regression function $f_0(x) := E[Y|X = x]$. The specific graph-based method we study is \emph{Laplacian eigenmaps}, first introduced by \cite{belkin03a}, which projects the response vector ${\bf Y} = (Y_1,\ldots,Y_n)$ onto the span of eigenvectors of a graph Laplacian. We focus on the the unnormalized graph Laplacian $L$, which is a difference operator acting on functions $f: \{X_1,\ldots,X_n\} \to \Reals$ as follows,
\begin{equation}
\label{eqn:graph_laplacian}
Lf(X_i) = \sum_{j = 1}^{n} \bigl(f(X_i) - f(X_j)\bigr)W_{ij}. 
\end{equation}
The Laplacian $L$ can be seen as a discretization of a weighted Laplace-Beltrami operator $\Delta_P$, defined when $P$ admits a density $p$ as
\begin{equation}
\label{eqn:fokker_planck_1}
\Delta_Pf= -\frac{1}{p} \mathrm{div}(p^2 \nabla f).
\end{equation}
The eigenvectors of $L$ form an orthonormal basis of $L^2(P_n)$, and are approximations to the eigenfunctions of $\Delta_P$. Their corresponding eigenvalues are likewise estimates of eigenvalues of $\Delta_P$, and thus give a notion of smoothness to each eigenvector---roughly speaking, the smaller the eigenvalue, the smoother the corresponding eigenvector.

Laplacian eigenmaps is thus a type \emph{spectral series} method. Spectral series methods, and more general orthogonal series methods, are some of most classical methods for non-parametric regression. Traditionally, orthogonal series methods are defined with respect to a fixed set of basis functions, orthogonalized in $L^2(Q)$ with respect to some pre-determined reference measure $Q$. Spectral series estimators take a particular orthogonal basis, the eigenfunctions of $\Delta_Q$. In contrast, eigenvectors of the graph Laplacian are data-dependent objects, and adapt to the geometry of the unknown design distribution $P$ in a rich manner.

The theoretical properties of (classical) spectral series methods are various, and are in general extremely well-understood \textcolor{red}{(TODO): references}. In particular, they are minimax rate optimal for nonparametric regression over H\"{o}lder and Sobolev spaces. However, these methods suffer from some serious drawbacks. As a practical matter, even computing the eigenfunctions of $\Delta_Q$ is in general non-trivial. Moreover, the statistical optimality of such estimators also requires that the reference measure $Q$ be equal to $P$, whereas in this work we shall always assume $P$ is unknown. Of course there exist workarounds for these issues, for instance non-parametric least squares or \textcolor{red}{(...)}. We will comment more on these alternatives later; for now we merely emphasize that they constitute a fundamental change to the method under consideration. \textcolor{red}{Last sentence is weak, need to clarify, can maybe discuss with Ryan + Siva.}

In contrast, Laplacian eigenmaps seeks to directly approximate a orthogonal series method that uses the eigenfunctions of $\Delta_P$, as we elaborate on momentarily. Importantly, it avoids the (arguably unrealistic) assumption that $P$ and the eigenfunctions of $\Delta_P$ be known, by using the graph Laplacian eigenvectors as a stand-in. On the other hand its theoretical statistical properties, and in particular its optimality as a method for regression, remain poorly understood.

@@

\begin{quote}
	Is Laplacian eigenmaps an optimal method for nonparametric regression over Sobolev functions?
\end{quote}

Broadly seeking, our answer is yes. We consider two different data models, one of which assumes a full-dimensional, flat Euclidean domain, and the other of which assumes that the distribution $P$ is supported on a manifold of small intrinsic dimension $m$. We show that when the regression function $f_0$ is smooth, in the Sobolev sense of having weak derivatives bounded in $\Leb^2$ norm, Laplacian eigenmaps methods are statistically minimax optimal, for both estimation and goodness-of-fit testing. Our statements holds for different relations between the dimension $d$ (or $m$) and number of derivatives $s$, depending on the problem (estimation or testing), and are summarized in Tables~\ref{tbl:estimation_rates} and~\ref{tbl:testing_rates}.
\begin{table}
	\begin{center}
		\begin{tabular}{p{.2\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Smoothness order & Flat Euclidean (Model~\ref{def:model_flat_euclidean}) & Manifold (Model~\ref{def:model_manifold}) \\
			\hline
			$s \leq 3$ & $\bm{n^{-2s/(2s + d)}}$ & $\bm{n^{-2s/(2s + m)}}$ \\
			$s > 3$  & $\bm{n^{-2s/(2s + d)}}$ & $n^{-6/(6 + m)}$
		\end{tabular}
	\end{center}
	\caption{Summary of estimation rates over Sobolev balls. Bold font marks minimax optimal rates. In each case, rates hold for all $d \in \mathbb{N}$ (under Model~\ref{def:model_flat_euclidean}), and for all $m \in \mathbb{N}, 1 < m < d$ (under Model~\ref{def:model_manifold}). Although we suppress it for simplicity, in all cases when the Laplacian eigenmaps estimator is optimal, the dependence of the error rate on the radius $M$ of the Sobolev ball is also optimal, as long as $M \gtrsim n^{-1/2}$.}
	\label{tbl:estimation_rates}
\end{table}

\begin{table}
	\begin{center}
		\begin{tabular}{p{.175\textwidth} p{.175\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Smoothness order & Dimension & Flat Euclidean (Model~\ref{def:model_flat_euclidean}) & Manifold (Model~\ref{def:model_manifold}) \\
			\hline
			\multirow{2}{*}{$s = 1$} & $\dim(\mc{X}) < 4$ & $\bm{n^{-4s/(4s + d)}}$ & $\bm{n^{-4s/(4s + m)}}$ \\
			& $\dim(\mc{X}) \geq 4$ & $\bm{n^{-1/2}}$ & $\bm{n^{-1/2}}$ \\
			\hline
			\multirow{3}{*}{$s = 2$ or $3$} & $\dim(\mc{X}) \leq 4$  & $\bm{n^{-4s/(4s + d)}}$ & $\bm{n^{-4s/(4s + m)}}$ \\
			& $4 <\dim(\mc{X}) < 4s$  & $n^{-2s/(2(s - 1) + d)}$ & $n^{-2s/(2(s - 1) + m)}$\\
			& $\dim(\mc{X}) \geq 4s$ & $\bm{n^{-1/2}}$ & $\bm{n^{-1/2}}$ \\
			\hline
			\multirow{3}{*}{$s > 3$} & $\dim(\mc{X}) \leq 4$ & $\bm{n^{-4s/(4s + d)}}$ & $n^{-12/(12 + d)}$ \\
			& $4 < \dim(\mc{X}) < 4s$ & $n^{-2s/(2(s - 1) + d)}$ & $n^{-6/(4 + m)}$ \\
			& $\dim(\mc{X}) \geq 4s$ & $\bm{n^{-1/2}}$ & $\bm{n^{-1/2}}$ \\
		\end{tabular}
	\end{center}
	\caption{Summary of Laplacian eigenmaps testing rates over Sobolev balls. Bold font marks minimax optimal rates. Rates when $d > 4s$ assume that $f_0 \in L^4(P,M)$. Although we suppress it for simplicity, in all cases when the Laplacian eigenmaps test is optimal, the dependence of the error rate on the radius $M$ of the Sobolev ball is also optimal, as long as $M \gtrsim n^{-1/2}$.}
	\label{tbl:testing_rates}
\end{table}

\paragraph{Related Work.}
\textcolor{red}{(TODO)}: Flesh this out.
\begin{itemize}
	\item Summarize what is known on fixed graphs, drawing from our paper on Laplacian smoothing.
	\item \textcolor{red}{von Luxburg, Belkin, Slepcev} show that for a fixed index $k$, the eigenvector $v_k$ converges to a eigenfunction $\psi_k$ of the Laplace-Beltrami operator $\Delta_P$. \textcolor{red}{Shi, Burago, Garcia Trillos, Calder, Cheng} build on this, giving finite sample bounds, rates of convergence, and making statements uniform over $1 \leq k \leq k_{\max}$. This makes it clear that Laplacian eigenmaps is approximating a classical spectral series method, and may be a good alternative when the latter cannot be implemented (e.g. because $P$ or $\mc{X}$ are unknown.) It is tempting to try to directly use these results along with the triangle inequality to deduce that if (i)  spectral series methods are optimal, and (ii) Laplacian eigenmaps is converging to a spectral series method, then (iii) Laplacian eigenmaps must be also optimal. Unfortunately, this won't work, both because we will need to take an increasing number of eigenvectors (increasing, that is, with the sample size $n$) to achieve an optimal estimator and test, and because the best known rates of convergence of $v_k \to \psi_k$ are slower than the minimax rates.
	\item There has been limited analysis of the estimator $\wh{f}$. \cite{zhou2011} consider $\wh{f}$, but in the semi-supervised setting where the number of unlabeled points grows to infinity. In this case $\wh{f}$ reduces to a classical spectral series estimator, and there is no error incurred by approximating $\Delta_P$. \cite{lee2016} consider the \emph{diffusion maps} estimator---which uses the eigenvectors of a different normalization of the Laplacian $L$---in both the supervised and semi-supervised setups, but they derive suboptimal rates for the supervised problem. As far as we know, there has been no analysis of the test statistic $\wh{T}$ in the random design framework which we study.
	\item Alternatively, \cite{trillos2020} \textcolor{blue}{(Green 2021)} use the graph Laplacian operator to induce a penalty over functions $f: \{\bf X\} \to \Reals$. The \emph{Laplacian smoothing} estimator $\wh{f}_{\mathrm{LS}}$ is obtained by minimizing the sum of this penalty along with a data-fidelity term,
	\begin{equation*}
	\wh{f}_{\mathrm{LS}} := \|Y - f\|_n^2 + \lambda \dotp{L_{n,\varepsilon}f}{f}_n.
	\end{equation*}
	\textcolor{blue}{(Green 2021)} show that the resulting estimator is minimax optimal, but only for $s = 1$ and $d \leq 4$. In contrast, the Laplacian eigenmaps estimator is optimal for all $s$ and $d$. 
\end{itemize}

\paragraph{Organization.}

We now outline the structure of the rest of this paper. In Section~\ref{sec:setup_main_results}, we formally define the regression problem and estimator we consider, and give some background about minimax regression over Sobolev spaces. In Sections~\ref{sec:minimax_optimal_laplacian_eigenmaps} and~\ref{sec:manifold_adaptivity}, we establish our main results regarding the optimality of Laplacian eigenmaps. The Laplacian eigenmaps estimator $\wh{f}$ is defined only at the design points $X_1,\ldots,X_n$. In Section~\ref{sec:out_of_sample}, we propose a method for out-of-sample extension of $\wh{f}$, and show that it has optimal out-of-sample error. In Section~\ref{sec:experiments} we examine the empirical behavior of Laplacian eigenmaps, and compare it to some natural competitors for non-parametric regression. We conclude with some discussion in Section~\ref{sec:discussion}.

\paragraph{Notation.}

@@

\begin{itemize}
	\item Define Lebesgue measure $\nu$. 
	\item Define Euclidean norm $\|\cdot\|$.
	\item Define $\Leb^q(\mc{X}), \Leb^2(P)$ and $\Leb^2(P_n)$. Define $C^k(\mc{X})$, and its norm. Define $C_0^k(\mc{X})$.
	\item Define asymptotic notation: $\ll$, $\lesssim$ and $\asymp$.
	\item Define min and max notation: $\vee$ and $\wedge$.
\end{itemize}

@@

In the following, we recall that the injectivity radius of a $m$-dimensional Riemannian manifold $\mc{X}$ is the maximum value of $\delta$ such that the exponential map $\exp_x: B_m(0,\delta) \subset T_x(\mc{X}) \to B_{\mc{X}}(x,\delta) \subset \mc{X}$ is a diffeomorphism for all $x \in \mc{X}$.
\begin{definition}[Manifold model]
	\label{def:model_manifold}
	The data $(X_1,Y_1),\ldots,(X_n,Y_n)$ are sampled according to~\eqref{eqn:model}. 
	The support $\mc{X}$ of the design distribution $P$ is a closed, connected, smooth and boundaryless Riemannanian manifold embedded in $\Rd$, of intrinsic dimension $1 \leq m < d$. The injectivity radius of $\mc{X}$ is upper bounded by a constant $i_0$. The design distribution $P$ admits a density $p$ with respect to the volume form $d\mu$ induced by the Riemannian structure of $\mc{X}$, which is bounded away from $0$ and $\infty$,
	\begin{equation*}
	0 < p_{\min} \leq p(x) \leq p_{\max} < \infty, \quad \textrm{for all $x \in \mc{X}$.}
	\end{equation*}
\end{definition}

\subsection{Laplacian eigenmaps}
We now formally define the estimator and test statistic we study. Both are derived from eigenvectors of the neighborhood graph Laplacian operator. For a positive, symmetric kernel $\eta: [0,\infty) \to [0,\infty)$, and a bandwidth parameter $\varepsilon > 0$, the \emph{neighborhood graph Laplacian} operator $L_{n,\varepsilon}$ is defined as
\begin{equation}
\label{eqn:neighborhood_graph_laplacian}
L_{n,\varepsilon}u(x) = \frac{1}{n\varepsilon^{2 + \mathrm{dim}(\mc{X})}} \sum_{i = 1}^{n} \bigl(u(x) - u(X_j)\bigr) \eta\biggl(\frac{\|x - X_j\|}{\varepsilon}\biggr)
\end{equation}
(Here $\mathrm{dim}(\mc{X})$ stands for the dimension of $\mc{X}$. It is equal to $d$ under the assumptions of Model~\ref{def:model_flat_euclidean}, and equal to $m$ under the assumptions of Model~\ref{def:model_manifold}. The pre-factor $(n\varepsilon^{2 + \mathrm{dim}(\mc{X})})^{-1}$ is purely for convenience in taking limits as $n \to \infty, \varepsilon \to 0$).

@@

% AG: Said elsewhere. 
% On a different note, it is worth explicitly pointing out that $\wh{f}$ is defined only in-sample, that is, only at the design points $X_1,\ldots,X_n$. There is no straightforward or agreed upon way to extend $\wh{f}$ out-of-sample, and we address the problem of out-of-sample estimation in some detail later in Section~\ref{sec:out_of_sample}. %

@@

\paragraph{Flat Euclidean case.}
Take $\mc{X}$ to be an open, connected set with Lipschitz boundary, as in Model~\ref{def:model_flat_euclidean}. Recall that for a given multiindex $\alpha \in \mathbb{N}^d$, a function $f$ is \emph{$\alpha$-weakly differentiable} if there exists some $h \in L^1(\mc{X})$ such that
\begin{equation*}
\int_{\mc{X}} h g = (-1)^{|\alpha|} \int_{\mc{X}} f D^{\alpha}g, \quad \textrm{for every $g \in C_0^{\infty}(\mc{X})$.}
\end{equation*}
If such a function $h$ exists, it is the $\alpha$th weak partial derivative of $f$, and denoted by $D^{\alpha}f = h$. If $f$ is $\alpha$-weakly differentiable for all $|\alpha| = j$, then $d^jf := \sum_{|\alpha| = j}D^{\alpha}f$ is the sum of all weak partial derivatives of order $j$.

For an integer $s \geq 1$, the Sobolev class $H^s(\mc{X})$ consists of all those functions $f \in \Leb^2(\mc{X})$ which have weak derivatives of all orders $j = 1,\ldots,s$, all of which also belong to $\Leb^2(\mc{X})$. The $j$th order semi-norm for $f \in H^s(\mc{X})$ is $|f|_{H^j(\mc{X})} := \|d^jf\|_{L^2(Q)}$, and the corresponding norm
\begin{equation*}
\|f\|_{H^s(\mc{X})}^2 := \|f\|_{\Leb^2(Q)}^2 + \sum_{j = 1}^{s} |f|_{H^j(Q)}^2,
\end{equation*}
induces a ball
\begin{equation*}
H^s(\mc{X};M) := \bigl\{f \in H^s(\mc{X}): \|f\|_{H^s(\mc{X})} \leq M\bigr\}.
\end{equation*} 
We note that $H^s(\mc{X})$ is the completion of $C^{\infty}(\mc{X})$ with respect to the $\|\cdot\|_{H^s(\mc{X})}$ norm, so that $C^{\infty}(\mc{X})$ is dense in $H^s(\mc{X})$.

\paragraph{Manifold case.}
There are several ways to define Sobolev spaces on a manifold \textcolor{red}{(references)}, but we will stick with a simple definition that suffices when $\mc{X}$ is a smooth Riemmanian manifold without boundary, as in Model~\ref{def:model_manifold}. Let $\{\phi_{\beta}, U_{\beta}\}$ be a $C^{\infty}$ atlas of $\mc{X}$, and let $\{h_\beta\}$ be a partition of unity subordinate to $\{(\phi_{\beta}), U_{\beta}\}$. For a function $f \in C^{\infty}(\mc{X})$, define the Sobolev norm to be,
\begin{equation*}
\|f\|_{H^s(\mc{X})}^2 = \sum_{\beta} \|(fh_{\beta}) \circ \phi_\beta^{-1}\|_{H^{k}(\Reals^m)}^2.
\end{equation*}
Then the order-$s$ Sobolev space $H^s(\mc{X})$ is defined to be the completion of $C^{\infty}(\mc{X})$ with respect to the norm $\|\cdot\|_{H^s(\mc{X})}$.

\paragraph{Boundary conditions.}
In the flat Euclidean case, when $s > 1$ we will assume that $f_0$ satisfies some boundary conditions. In particular, we will assume that $f_0$ belongs to the set of \emph{zero trace} Sobolev functions $H_0^{s}(P)$, defined as the closure of $C_0^s(\mc{X})$ with respect to $\|\cdot\|_{H^s(\mc{X})}$ norm. Of course, under Model~\ref{def:model_manifold} the domain $\mc{X}$ is without boundary, and so boundary conditions are irrelevant.
